```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
pacman::p_load(tidyverse, janitor)
load("data/data.RData")
source(here::here("ggplot_theme.r"))
```

Check variables distributions 

```{r}
pacman::p_load(ggstance) # for stat_summaryh
```

```{r}
dat %>% 
  pivot_longer(cols = c("DSI", "global_inc"), 
               names_to = "var", 
               values_to = "val") %>%    
  mutate(harvest_year=forcats::fct_rev(factor(harvest_year))) %>%
  mutate(var=fct_recode(var, 
                        `Smut severity index` = "DSI", 
                        `Smut incidence` = "global_inc")) %>%
  ggplot() + 
  aes(val)+
  geom_histogram(alpha=.6, fill="steelblue")+
  facet_grid(harvest_year~var, switch="x", scales = "free_x")+
  stat_summaryh(fun.x=median,
                geom="vline", alpha=.2,
                aes(xintercept=..x.., 
                y=0, 
                    group = interaction(var, harvest_year)),
                size=.51, colour="red") +
  stat_summaryh(fun.x=median,
                geom="text",
                aes(y=10, 
                    label=..x.. %>% round(1)),
                size=3, colour="red", angle=0, vjust=-.1) +
  stat_summaryh(fun.x=max,
                geom="vline", alpha=.2,
                aes(xintercept=..x.., 
                y=0, 
                    group = interaction(var, harvest_year)),
                size=.51, colour="grey") +
  stat_summaryh(fun.x=max,
                geom="text",
                aes(y=10, label=..x.. %>% round(1)),
                size=3, colour="grey", angle=0, vjust=-.1) +
  # ylim(0,15)+
  labs(x="", y = "Fields count")+
  theme_bw() + 
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_line(size=.1), 
        strip.background = element_blank())+
  theme(strip.placement = "outside")
```

```{r, eval=FALSE}
ggsave(last_plot(), 
       file = "plots/histograms.png", 
       width = 5, height = 4)
```

Disease class frequency along the years

```{r}
dat %>% 
  pivot_longer(cols = starts_with("class_"), 
               names_to = "var", 
               values_to = "val") %>%   
  group_by(harvest_year,var) %>% 
  summarise(sum_n=sum(val)) %>% 
  group_by(harvest_year) %>% 
  mutate(freq = sum_n / sum(sum_n)*100, 
         var=forcats::fct_rev(var), 
         harvest_year = factor(harvest_year)) %>%
  # filter(!harvest_year %in% c(2015,2019,2020)) %>% 
  ggplot(aes(
    x = harvest_year,
    # x = reorder(Zona, super, sum),
    y = freq, 
    fill = var)) + 
  geom_bar(stat = "identity", color = gray(.5)) +
  geom_text(aes(label = freq %>% round(1)), 
            position = position_stack(vjust = 0.5),
            col = "white", fontface = "bold", size = 3)+
  scale_fill_viridis_d(begin = .1, end =.9, direction = 1)+
  labs(fill= "Disease\nseverity class", 
       x="Harvest year", 
       y = "Frequency (%)")+
  theme_bw()
```

```{r, eval=FALSE}
ggsave(last_plot(), 
       file = "plots/stacked_bar.png", 
       width = 5, height = 5)
```

```{r}
dat %>% 
  ggplot()+
  aes(x=DSI, 
      y=global_inc)+
  geom_point()+
  labs(x= "DSI",
       y="Disease incidence (%)")+
  my_theme + conc_set 
```

```{r, eval=FALSE}
ggsave(last_plot(), 
       file = "plots/DSI_inc.png", 
       width = 5, height = 5)
```

Let's have a first visualization of the data. As a first step, we wanted to see the relationship between each severity class incidence vs the field incidence, then each point represents one field comparison of the pair x-class incidence vs field incidence. The closest the point is to the 1:1 diagonal line, the better it explain the field incidence. 

```{r}
dat %>% 
  pivot_longer(
    cols = starts_with("class."),
    names_to = "metric", 
    values_to = "Severity class incidence") %>% 
  mutate(metric = str_replace(metric, "_inc", "")) %>% 
  # count(metric)
  ggplot()+
  aes(x=global_inc, 
      y=`Severity class incidence`)+
  geom_point()+
  labs(x="Disease incidence")+
  facet_wrap("metric")+
  my_theme + conc_set 
```

As we adopted a mixed regression model approach to estimate the regression coefficients based on the id-level regressions, we kept only those environments (department * year) with more than 3 fields assessed and at least 5% of peanut smut incidence range. This restriction allowed us to estimate an id-level regression as random effect and the overall line regression as fixed effect.  

```{r}
dat %>%   
  group_by(harvest_year, department)  %>% 
  mutate(min = min(global_inc),
         max = max(global_inc),         
         range = max-min,
         n = n(), 
         keep = case_when(range >4.99 & n > 3 ~ "yes", 
                          TRUE ~ "no")) %>%   
  ggplot()+
  aes(y=id, x=global_inc, col=keep)+
  geom_pointrangeh(aes(xmin = min, xmax = max), size=.1, alpha=.7)+
  my_theme2+
  labs(x="Peanut smut incidence")
               
```

Let's call "dat1" to the ready-to-model dataset

```{r}
dat1 <- dat %>% 
  group_by(harvest_year, department)  %>% 
  mutate(min = min(global_inc),
         max = max(global_inc),         
         range = max-min,
         n = n(), 
         keep = case_when(range >4.99 & n > 3 ~ "yes", 
                          TRUE ~ "no")) %>%   
  ungroup %>% 
  filter(keep=="yes") %>% 
  as.data.frame() # for avoiding bug with ggpredict package
```

Check dat1 

```{r}
dat1 %>% 
  tabyl(department, harvest_year) %>% 
  adorn_totals()
```

Have a glance of dat1

```{r}
dat1 %>% head(15)
```

# Modeling

```{r}
pacman::p_load(car, 
               nlme,        # for adjusting the mixed linear regression 
               lattice, 
               performance, # for calculating model R2
               ggeffects,   # for estimating predicted values
               ggtext,      # for including model coefficients and R2 in the plots
               glue         # for writing model coefficients and R2
               )
ctrl <- lmeControl(opt='optim')
```

Class 1 incidence vs Disease incidence

As most of the field disease assessments we have lots of the observations accumulated in the <10% incidence portion. That could lead to heteroscedasticity issues in our regression fit. So we used the "lme" function from "nlme" package to deal with the increased variance along the x-axis. We included the argument `weights=varPower(form = ~global_inc)` for modelling the variance.  
   

```{r}
m1<-lme(class.1_inc ~ global_inc, 
          random = ~global_inc|id,
          weights=varPower(form = ~global_inc),
          data=dat1)
```


```{r, eval=F}
png(filename="plots/class1_reg_diagnostics.png",
    type="cairo",
    units="in",
    width=4,
    height=4,
    pointsize=12,
    res=96)
```


```{r}
plot(m1)
dev.off()
```

```{r}
summary(m1)
```

Id-level plots

```{r}
dat1 %>% 
  ggplot() +
  aes(x = global_inc, y = class.1_inc)+
  facet_wrap(~id) +
  geom_point(alpha = 0.5) +
  geom_line(data = cbind(dat1, pred = predict(m1)), aes(y = pred), size = 1)+
  my_theme2

# ggsave(last_plot(), file = "plots/class1_id_level.png", units="in", w=6, h=4)
```

Fixed effects plot 

```{r}
df1 <- data.frame(x = 0, y = 50,
  label = glue(
    "y = {round(summary(m1)$tTable[1,1],3)} + {round(summary(m1)$tTable[2,1],3)} * x\n"
    )  
  ) 

p1 <- ggpredict(m1, terms="global_inc [all]") %>% 
  ggplot()+
  aes(x = x, y = predicted) +
  geom_smooth(data=dat1, aes(x = global_inc, y = class.1_inc, group=id), 
              method="lm", se=F, col="grey30", size=.5)+
  geom_point(data=dat1, aes(x = global_inc, y = class.1_inc), size=.3)+
  geom_line(size=1, col="#3366FF")+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) +
  labs(y="Class-1 Incidence ",  x = element_blank())+
  my_theme + conc_set +
  geom_richtext(
    data = df1,
    aes(x=x, y=y, label = label),
    hjust = 0, vjust = 1, 
    label.color = NA,
    # fill = alpha("grey30", .2)
  )
p1
```

Continue with all the class levels...

```{r}
m2<-lme(class.2_inc~global_inc, 
          random = ~global_inc|id,
          weights=varPower(form = ~global_inc),
          data=dat1)
```

```{r}
png(filename="plots/class2_reg_diagnostics.png",
    type="cairo",
    units="in",
    width=4,
    height=4,
    pointsize=12,
    res=96)
```

```{r}
plot(m2)
```

```{r, eval=F}
dev.off()
```

```{r}
summary(m2)
```

```{r}
dat1 %>% 
  ggplot() +
  aes(x = global_inc, y = class.2_inc)+
  facet_wrap(~id) +
  geom_point(alpha = 0.5) +
  geom_line(data = cbind(dat1, pred = predict(m2)), aes(y = pred), size = 1)+
  my_theme2

ggsave(last_plot(), file = "plots/class2_id_level.png", units="in", w=6, h=4)
```

```{r}
df2 <- data.frame(x = 0, y = 50,
  label = glue(
    "y = {round(summary(m2)$tTable[1,1],2)} + {round(summary(m2)$tTable[2,1],3)} * x\n"
    # (*R<sup>2</sup>* = {round(r2_nakagawa(m2)$R2_conditional, 3)})"
    )  
  ) 

p2 <- ggpredict(m2, terms="global_inc [all]") %>% 
  ggplot() +
  aes(x = x, y = predicted)+
  geom_smooth(data=dat1, aes(x = global_inc, y = class.2_inc, group=id), 
              method="lm", se=F, col="grey30", size=.5, alpha=.5)+
  geom_point(data=dat1, aes(x = global_inc, y = class.2_inc), size=.3)+
  geom_line(size=1, col="#3366FF")+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) +
  labs(y="Class-2 Incidence",  x = element_blank())+
  my_theme+conc_set + 
    geom_richtext(
    data = df2,
    aes(x=x, y=y, label = label),
    hjust = 0, vjust = 1, 
    label.color = NA
    )
p2
```


```{r}
m3<-lme(class.3_inc~global_inc, 
        random = ~global_inc|id,
        control=ctrl,
        weights=varPower(form = ~global_inc),
        data=dat1)
```


```{r, eval=F}
png(filename="plots/class3_reg_diagnostics.png",
    type="cairo",
    units="in",
    width=4,
    height=4,
    pointsize=12,
    res=96)
```


```{r}
plot(m3)
```


```{r, eval=F}
dev.off()
```

```{r}
dat1 %>% 
  ggplot() +
  aes(x = global_inc, y = class.3_inc)+
  facet_wrap(~id) +
  geom_point(alpha = 0.5) +
  geom_line(data = cbind(dat1, pred = predict(m3)), aes(y = pred), size = 1)+
  my_theme2

ggsave(last_plot(), file = "plots/class3_id_level.png", units="in", w=6, h=4)
```

```{r}
df3 <- data.frame(x = 0, y = 50,
  label = glue(
    "y = {round(summary(m3)$tTable[1,1],3)} + {round(summary(m3)$tTable[2,1],3)} * x\n")  
  ) 

p3 <- ggpredict(m3, terms="global_inc [all]") %>% 
  ggplot() +
  aes(x = x, y = predicted)+
  geom_smooth(data=dat1, aes(x = global_inc, y = class.3_inc, group=id), 
              method="lm", se=F, col="grey30", size=.5)+
  geom_point(data=dat1, aes(x = global_inc, y = class.3_inc), size=.3)+
  geom_line(size=1, col="#3366FF")+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) +
  labs(y="Class-3 Incidence ",  x = element_blank())+
  my_theme+conc_set + 
  geom_richtext(
    data = df3,
    aes(x=x, y=y, label = label),
    hjust = 0, vjust = 1, 
    label.color = NA)
p3
```

```{r}
m4<-lme(class.4_inc~global_inc, 
          random = ~global_inc|id,
          control=ctrl,
          weights=varPower(form = ~global_inc),
          data=dat1)
```


```{r}
png(filename="plots/class4_reg_diagnostics.png",
    type="cairo",
    units="in",
    width=4,
    height=4,
    pointsize=12,
    res=96)
```


```{r}
plot(m4)
```


```{r}
dev.off()
```

```{r}
dat1 %>% 
  ggplot() +
  aes(x = global_inc, y = class.4_inc)+
  facet_wrap(~id) +
  geom_point(alpha = 0.5) +
  geom_line(data = cbind(dat1, pred = predict(m4)), aes(y = pred), size = 1)+
  my_theme2

ggsave(last_plot(), file = "plots/class4_id_level.png", units="in", w=6, h=4)
```


```{r}
df4 <- data.frame(x = 0, y = 50,
  label = glue(
    "y = {round(summary(m4)$tTable[1,1],3)} + {round(summary(m4)$tTable[2,1],3)} * x\n")  
  ) 

p4 <- ggpredict(m4, terms="global_inc [all]") %>% 
ggplot() +
  aes(x = x, y = predicted)+
  geom_smooth(data=dat1, aes(x = global_inc, y = class.4_inc, group=id), 
              method="lm", se=F, col="grey30", size=.5)+
  geom_point(data=dat1, aes(x = global_inc, y = class.4_inc), size=.3)+
  geom_line(size=1, col="#3366FF")+    
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) +
  labs(y="Class-4 Incidence",  x = element_blank())+
  my_theme + 
  conc_set + 
    geom_richtext(
    data = df4,
    aes(x=x, y=y, label = label),
    hjust = 0, vjust = 1, 
    label.color = NA) 
p4
```

```{r}
pacman::p_load(patchwork)
p.1234 <- (p1 + p2) / (p3 + p4) + 
  plot_annotation(
    caption = 'Disease incidence',
    theme = theme(plot.caption = element_text(hjust = .5, size = 12))
  )
p.1234
```

As we can see, class 3 and 4 regressions are the most close to the diagonal line. So we considered including a class 3 + class 4 incidence, to test against the disease incidence.  

```{r}
dat1<- dat1 %>% 
  mutate(class.3.4_inc=(class_3+class_4)/total_pods*100)

m3.4<-lme(class.3.4_inc~global_inc, 
          random = ~global_inc|id,
          control=ctrl,
          weights=varPower(form = ~global_inc),
          data=dat1)
```


```{r}
png(filename="plots/class3.4_reg_diagnostics.png",
    type="cairo",
    units="in",
    width=4,
    height=4,
    pointsize=12,
    res=96)
```


```{r}
plot(m3.4)
dev.off()
```

```{r}
dat1 %>% 
  ggplot() +
  aes(x = global_inc, y = class.3.4_inc)+
  facet_wrap(~id) +
  geom_point(alpha = 0.5) +
  geom_line(data = cbind(dat1, pred = predict(m3.4)), aes(y = pred), size = 1)+
  my_theme2

ggsave(last_plot(), file = "plots/class3.4_id_level.png", units="in", w=6, h=4)
```

```{r}
df3.4 <- data.frame(x = 0, y = 50,
  label = glue(
    "y = {round(summary(m3.4)$tTable[1,1],3)} + {round(summary(m3.4)$tTable[2,1],3)} * x\n")  
  ) 

p3.4 <- ggpredict(m3.4, terms="global_inc [all]") %>% 
  ggplot() +
  aes(x = x, y = predicted)  +
  geom_smooth(data=dat1, aes(x = global_inc, y = class.3.4_inc, group=id), 
              method="lm", se=F, col="grey30", size=.5)+
  geom_point(data=dat1, aes(x = global_inc, y = class.3.4_inc), size=.3)+
  geom_line(size=1, col="#3366FF")+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) +
  labs(y="Class-3+4 Incidence ", x="")+
  my_theme + 
  conc_set + 
  geom_richtext(
    data = df3.4,
    aes(x=x, y=y, label = label),
    hjust = 0, vjust = 1, 
    label.color = NA
  ) 
p3.4
```

```{r}
(p.1234 | p3.4 )+
    plot_annotation(
    caption = 'Disease incidence',
    theme = theme(plot.caption = element_text(hjust = .5, size = 12))
  )
```

```{r, eval =F}
ggsave(last_plot(), file = "plots/regressions.png", width = 9, height = 6)
```
